# Audit: impact of the `get_ll` precedence fix {.unnumbered}

::: {.callout-important}
Internal audit document. This report quantifies the effect of correcting an
operator-precedence error in the legacy likelihood function on the Bayes
factor results reported in the 2020 manuscript drafts ("Running to a
standstill"). It exists so that any circulated or submitted numbers can be
traced and, where necessary, corrected.
:::

## The defect

The legacy log-likelihood (`scripts/load_packages_and_functions.R`, line 13,
preserved on branch `old/pre-refactor`) was written:

```r
- n * log(sig) - (n/2) * log(2 * pi) - (1 / 2 * sig_sq) * sum((x - mu)^2)
```

R parses `(1 / 2 * sig_sq)` as $\frac{1}{2}\sigma^2$, whereas the normal
log-likelihood requires $\frac{1}{2\sigma^2}$. The first two terms are
scenario-invariant and cancel in any likelihood ratio, so the Bayes factor
is affected only through the quadratic term:

$$
\log \mathrm{BF}_{\text{legacy}}
 = \frac{\sigma^2}{2}\left(S_0 - S_\mu\right), \qquad
\log \mathrm{BF}_{\text{correct}}
 = \frac{1}{2\sigma^2}\left(S_0 - S_\mu\right),
$$

where $S_\mu = \sum_i (x_i - \mu)^2$. Hence

$$
\log \mathrm{BF}_{\text{legacy}} = \sigma^4 \, \log \mathrm{BF}_{\text{correct}}.
$$

Two consequences follow immediately, before any recomputation:

1. **The scenario that maximises the BF is unchanged** --- both versions are
   strictly increasing transformations of $S_0 - S_\mu$ --- so every
   "slowdown of x% is most likely" statement in the manuscript is
   unaffected.
2. **Every reported BF magnitude is wrong.** The standard deviation of
   annual $e_0$ change is roughly 0.1--0.25 years across UK populations, so
   $\sigma^4 \approx 10^{-4}$--$10^{-3}$: legacy log-BFs are three to four
   orders of magnitude too small, which is why the manuscript's Table 4
   reports values such as "61% (BF: 1.00135)" --- BFs indistinguishable
   from 1 despite eight years of accumulated evidence.

## Recomputation on the original data window

Both versions are re-run below on exactly the manuscript's setup: actual
$e_0$ from the ONS single-year life tables (`data/e0_from_ons_allnations.csv`,
1980--2018), calibration on annual changes 1991--2010, observation windows
2011--2011 through 2011--2018, scenarios 0--100% slowdown in 1% steps.

```{r}
#| label: recompute
pacman::p_load(tidyverse, here, knitr)
source(here("R", "bayes_factors.R"))

dta_e0 <- read_csv(here("data", "e0_from_ons_allnations.csv"),
                   show_col_types = FALSE)

populations <- c("United Kingdom", "England", "Scotland", "Wales",
                 "Northern Ireland")

ch_data <- dta_e0 %>%
  filter(population %in% populations) %>%
  group_by(population, sex) %>%
  arrange(year) %>%
  mutate(ch_e0 = e0 - lag(e0)) %>%
  ungroup() %>%
  filter(!is.na(ch_e0))

run_both <- function(dta, after_end) {
  bind_rows(
    calc_bayes_factors(dta, c(1991, 2010), c(2011, after_end), ch_e0,
                       ll_fun = get_ll_legacy) %>%
      mutate(version = "legacy"),
    calc_bayes_factors(dta, c(1991, 2010), c(2011, after_end), ch_e0,
                       ll_fun = get_ll) %>%
      mutate(version = "corrected")
  )
}

bf_both <- ch_data %>%
  group_by(population, sex) %>%
  nest() %>%
  crossing(after_end = 2011:2018) %>%
  mutate(bf = map2(data, after_end, run_both)) %>%
  select(population, sex, after_end, bf) %>%
  unnest(bf)
```

### Does the maximising scenario ever differ?

```{r}
#| label: argmax-check
argmax_tbl <- bf_both %>%
  group_by(population, sex, after_end, version) %>%
  slice_max(log_bf, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(population, sex, after_end, version, perc) %>%
  pivot_wider(names_from = version, values_from = perc)

stopifnot(all(argmax_tbl$legacy == argmax_tbl$corrected))
n_cells <- nrow(argmax_tbl)
```

Across all `r n_cells` population-sex-window combinations the BF-maximising
slowdown percentage is **identical** under the legacy and corrected
likelihoods, confirming the algebraic argument above.

### Corrected magnitudes for the full 2011--2018 window

```{r}
#| label: headline-table
headline <- bf_both %>%
  filter(after_end == 2018) %>%
  group_by(population, sex, version) %>%
  slice_max(log_bf, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(population, sex, version,
            slowdown_pct = 100 * (1 - perc),
            bf = bayes_factor, log10_bf = log_bf / log(10)) %>%
  pivot_wider(names_from = version, values_from = c(bf, log10_bf)) %>%
  select(population, sex, slowdown_pct,
         bf_legacy, bf_corrected, log10_bf_corrected) %>%
  arrange(match(population, populations), sex)

headline %>%
  kable(
    digits = c(0, 0, 0, 5, 1, 1),
    col.names = c("Population", "Sex", "BF-maximising slowdown (%)",
                  "BF (legacy, as reported)", "BF (corrected)",
                  "log10 BF (corrected)"),
    caption = "Maximised Bayes factors, 2011-2018 window, legacy vs corrected likelihood."
  )
```

The legacy column reproduces the manuscript's Table 4 values (e.g. UK "61%
(BF: 1.001...)"), and the maximising percentages match Table 4 exactly
(61% UK both sexes; 59%/73% Scotland m/f; 83%/73% Wales m/f; 15% Northern
Ireland males), confirming provenance. The corrected column is what should
have been reported. On the @kass1995bayes scale the correction moves males
in the UK and England from "not worth more than a bare mention" (an
artefact of the bug) to "strong" evidence for a slowdown
($\mathrm{BF} \approx 70$--$100$), females to "positive" evidence
($\mathrm{BF} \approx 2$--$4$), while Northern Ireland males remain
genuinely equivocal ($\mathrm{BF} \approx 1.1$). Notably, under the
corrected likelihood the evidence for a slowdown is *stronger for males
than females* --- the draft manuscript's suggestion that ONS projections
"may still be underestimating the stalling for females" rested partly on
the attenuated magnitudes and deserves re-examination in the update.

### Evidence accumulation, corrected

```{r}
#| label: fig-accumulation
#| fig-width: 8
#| fig-height: 5
bf_both %>%
  filter(population == "United Kingdom") %>%
  group_by(sex, after_end, version) %>%
  slice_max(log_bf, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  ggplot(aes(x = after_end, y = log_bf / log(10),
             colour = sex, linetype = version)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Final year of observation window (from 2011)",
    y = "log10 Bayes factor at maximising scenario",
    title = "United Kingdom: maximised Bayes factor as evidence accumulates",
    subtitle = "Legacy values are visually indistinguishable from zero at this scale"
  )
```

## Affected outputs

| Output | Status |
|---|---|
| Manuscript Table 4 (`bayes_paper/tables/M04_bf_nochange.xlsx`, `bf_nochange.xlsx`) | BF magnitudes wrong; maximising percentages correct |
| Manuscript Table 5 (`bayes_paper/tables/M05_bf_cf_ons_projections.xlsx`) | BF magnitudes wrong; implied-slowdown percentages correct |
| Figure `A12_Bayes_Factor_Schedules.png` | y-axis values wrong (shape and argmax correct) |
| Figures `M03_ons_cf_bayes.png`, `bayes_factor_2018_update.png`, `bayes_factor_various_proposed_slowdown_levels.png` | as above where BF values are shown |
| Abstract / results prose quoting BF values | wrong wherever a BF magnitude is quoted; "most likely slowdown" percentages unaffected |

## Recommendations

1. If any version of the manuscript containing BF magnitudes was submitted,
   circulated, or cited, issue a correction note alongside any reissue;
   the qualitative conclusions (existence and approximate magnitude of the
   slowdown) stand.
2. All updated analyses (2011--2024 windows) use the corrected `get_ll`
   from `R/bayes_factors.R`; report `log10 BF` rather than raw BF, since
   corrected BFs overflow `exp()` for long windows.
3. A minor prose inconsistency found during this audit: the manuscript
   describes the calibration window variously as "1990 to 2010" and
   "1991-2010"; the code uses annual changes for years 1991--2010 (i.e.
   $e_0$ levels 1990--2010). Standardise on the code's definition.
