---
title: "Trying openxlsx with tables"
output: html_notebook
---

This document will try to write formatted tables as worksheets in excel using `openxlsx` and `kable`/`kableExtra`.

```{r}
pacman::p_load(openxlsx, tidyverse, kableExtra)

```

Let's use some default datasets

```{r}
data("ChickWeight")
chickwts <- as_tibble(chickwts)
```

```{r}

basic_table <- 
  chickwts %>% 
    group_by(feed) %>% 
    summarise(mean_wt = mean(weight), 
              sd_wt   = sd(weight),
              min_wt  = min(weight),
              max_wt  = max(weight),
              n_obs   = n()
    ) %>% 
    ungroup()

basic_table
```

Write out the basic table 


```{r}
wb <- openxlsx::createWorkbook()

wb %>% addWorksheet(sheetName = "basic_table")
wb %>% writeData(sheet = "basic_table", x = basic_table)

saveWorkbook(wb, file = here("bayes_paper", "tables", "test_workbook.xlsx"), overwrite = TRUE)

```

This works. Now to see what happens when the object created is a formatted table 

```{r}
fancy_table <- 
  basic_table %>% 
    kable() %>% 
    kable_styling() 
  
fancy_table

```

```{r}
wb <- openxlsx::createWorkbook()

wb %>% addWorksheet(sheetName = "basic_table")
wb %>% writeData(sheet = "basic_table", x = basic_table)

wb %>% addWorksheet(sheetName = "fancy_table")
wb %>% writeData(sheet = "fancy_table", x = fancy_table)


saveWorkbook(wb, file = here("bayes_paper", "tables", "test_workbook.xlsx"), overwrite = TRUE)

```

So, this doesn't work. Instead let's first look at `writeDataTable`, then at formatting efficiently 

```{r}
wb <- openxlsx::createWorkbook()

wb %>% addWorksheet(sheetName = "basic_table")
wb %>% writeData(sheet = "basic_table", x = basic_table)

wb %>% addWorksheet(sheetName = "basic_datatable")
wb %>% writeData(sheet = "basic_datatable", x = basic_table)


saveWorkbook(wb, file = here("bayes_paper", "tables", "test_workbook.xlsx"), overwrite = TRUE)


```


No obvious advantges to this, but can see the `mergeCells` option, so let's think about how to use that 

```{r}
make_fake_mtrx <- function(x){
  tmp <- matrix(nrow = 0, ncol = length(x))
  colnames(tmp) <- x
  tmp
}

wb <- openxlsx::createWorkbook()

wb %>% addWorksheet(sheetName = "basic_table")
wb %>% writeData(sheet = "basic_table", x = basic_table, startRow = 2)
wb %>% mergeCells(sheet = "basic_table", rows = 1, cols = 2:5)
wb %>% writeData(sheet = "basic_table", x = make_fake_mtrx(c(" ",rep("Weight", 4), "N")))


saveWorkbook(wb, file = here("bayes_paper", "tables", "test_workbook.xlsx"), overwrite = TRUE)



```

This approach isn't pretty, but it works...

