# Bayes factors versus ONS projection assumptions

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
pacman::p_load(tidyverse, here, readxl, knitr, kableExtra)

# Legacy helper first (non-BF utilities only); corrected engine LAST so its
# get_ll / calc_bayes_factors mask the buggy legacy definitions (see audit
# appendix and NOTES.md).
source(here::here("scripts", "load_packages_and_functions.R"))
source(here::here("R", "bayes_factors.R"))

populations <- c("United Kingdom", "England", "Scotland", "Wales", "Northern Ireland")

# Actual e0 to 2024 (single-year life tables to 2018, 2024-based NPP historic
# estimates 2019-2024). ONS projection rounds now cover 2012-2024.
dta_e0 <- read_csv(here("data", "e0_from_ons_allnations_extended.csv"),
                   show_col_types = FALSE) %>%
  filter(population %in% populations)

ons_projections_tidied <- read_rds(here("data", "ons_projections", "all_tidied_2012_2024.rds"))
```

This chapter compares the improvement rate that maximises the Bayes factor schedule for the **pre-pandemic headline window** (2011–2019, from the previous chapter) against the average annual improvement rate for $e_0$ implied by each ONS projection round (2012, 2014, 2016, 2018, 2020, 2022 and 2024) [@ons_expectation_of_life], to quantify how optimistic or pessimistic each projection has been relative to what the pre-2020 data alone implied.

The Bayes-factor benchmark is a single value per population and sex: the mean annual $e_0$ gain at the pre-pandemic BF-maximising slowdown scenario. Anchoring on the pre-pandemic window (rather than a series that includes the COVID shock) keeps the comparison against ONS assumptions clean — the ONS rounds themselves are, likewise, projections of a smoothly improving process.

```{r}
#| label: recalc-bayes-factors
#| warning: false
ch_e0_dta <-
  dta_e0 %>%
  group_by(population, sex) %>%
  arrange(year) %>%
  mutate(ch_e0 = e0 - lag(e0)) %>%
  ungroup() %>%
  filter(!is.na(ch_e0)) %>%
  select(-e0, -source)

# Pre-pandemic (2011-2019) BF-maximising mean annual gain per population / sex.
bf_benchmark <-
  ch_e0_dta %>%
  group_by(population, sex) %>%
  nest() %>%
  mutate(bf = map(data, ~calc_bayes_factors(
    .x, before_period = c(1991, 2010), after_period = c(2011, 2019),
    outcome_var = ch_e0))) %>%
  select(population, sex, bf) %>%
  unnest(bf) %>%
  group_by(population, sex) %>%
  slice_max(log_bf, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(population, sex, bf_mu = mu)

bf_benchmark
```

For each ONS projection round we compute the mean annual improvement in period life expectancy at birth implied by that round (from the projection's period $e_0$ series in the projection year onwards), and set it alongside the pre-pandemic Bayes-factor benchmark.

```{r}
#| label: tbl-bf-vs-ons
#| warning: false
#| message: false
ons_implied <-
  ons_projections_tidied %>%
  mutate(sex = case_when(
    sex == "f" ~ "female",
    sex == "m" ~ "male",
    TRUE       ~ NA_character_
  )) %>%
  filter(age == 0) %>%
  filter(year >= proj_year) %>%
  group_by(country, sex, proj_year) %>%
  arrange(year) %>%
  mutate(ch_e0 = ex - lag(ex)) %>%
  summarise(ons_mu = mean(ch_e0, na.rm = TRUE), .groups = "drop") %>%
  mutate(country = ifelse(country == "UK", "United Kingdom", country)) %>%
  rename(population = country, round = proj_year)

# Modern dplyr will not join a plain factor to an ordered factor, and mixing
# population key types silently drops rows: coerce both sides to character
# before the join (behaviour-preserving).
combined_long <-
  ons_implied %>%
  mutate(population = as.character(population)) %>%
  left_join(bf_benchmark %>% mutate(population = as.character(population)),
            by = c("population", "sex")) %>%
  mutate(population = factor(population, levels = populations)) %>%
  arrange(population, sex, round)

combined_wide <-
  combined_long %>%
  pivot_wider(names_from = sex, values_from = c(ons_mu, bf_mu)) %>%
  select(population, round,
         ons_mu_female, bf_mu_female, ons_mu_male, bf_mu_male) %>%
  arrange(population, round)

combined_wide %>%
  kable(
    digits = 3,
    col.names = c("Population", "Round", "ONS", "BF (pre-2020)", "ONS", "BF (pre-2020)")
  ) %>%
  add_header_above(c(" " = 2, "Female" = 2, "Male" = 2)) %>%
  kable_styling()
```

The Bayes-factor benchmark (`BF (pre-2020)`) is constant down each population block, because it is the single pre-pandemic best estimate; the ONS column changes with each round. A round whose implied gain lies below the benchmark was projecting slower improvement than the pre-2020 data alone implied (relatively pessimistic); above it, relatively optimistic.

Now a figure for the UK as a whole, showing every ONS round against the fixed pre-pandemic Bayes-factor benchmark.

```{r}
#| label: fig-ons-cf-bayes
#| warning: false
uk_long <-
  combined_long %>%
  filter(population == "United Kingdom") %>%
  select(round, sex, `ONS projection` = ons_mu, `Bayes Factor (pre-2020)` = bf_mu) %>%
  pivot_longer(c(`ONS projection`, `Bayes Factor (pre-2020)`),
               names_to = "method", values_to = "gain") %>%
  mutate(gain = 52.18 * gain)

uk_long %>%
  ggplot(aes(x = round, y = gain, colour = method, shape = method, linetype = method)) +
  geom_line() +
  geom_point() +
  facet_wrap(~sex) +
  expand_limits(y = 0) +
  labs(
    x = "ONS projection round", y = "Implied gain (weeks/year)",
    title = "ONS-implied vs pre-pandemic Bayes-factor annual e0 gain",
    subtitle = "UK as a whole; Bayes-factor benchmark is fixed (pre-pandemic 2011-2019)"
  )

ggsave(here("bayes_paper", "figures", "M03_ons_cf_bayes_2024.png"), height = 12, width = 24, units = "cm", dpi = 300)
```

```{r}
#| label: newest-rounds-values
#| echo: false
uk_cmp <-
  combined_long %>%
  filter(population == "United Kingdom") %>%
  mutate(ons_minus_bf = ons_mu - bf_mu)

uk_get <- function(sx, rd) uk_cmp %>% filter(sex == sx, round == rd) %>% pull(ons_mu)
bf_f <- bf_benchmark %>% filter(population == "United Kingdom", sex == "female") %>% pull(bf_mu)
bf_m <- bf_benchmark %>% filter(population == "United Kingdom", sex == "male") %>% pull(bf_mu)
```

Reading down the UK block, the ONS rounds tracked steadily downwards from the 2012 round, reflecting the accumulating evidence of stalling: for UK females the implied gain fell from `r formatC(uk_get("female", 2012), format="f", digits=3)` years/year (2012 round) toward the pre-pandemic Bayes-factor benchmark of `r formatC(bf_f, format="f", digits=3)`, and for males from `r formatC(uk_get("male", 2012), format="f", digits=3)` toward `r formatC(bf_m, format="f", digits=3)`.

The newer, post-drafting rounds (2020-based interim, 2022-based and 2024-based) sit close to the pre-2020 benchmark rather than far below it. For UK females the 2020, 2022 and 2024 rounds imply `r formatC(uk_get("female", 2020), format="f", digits=3)`, `r formatC(uk_get("female", 2022), format="f", digits=3)` and `r formatC(uk_get("female", 2024), format="f", digits=3)` years/year respectively — all at or slightly **above** the pre-2020 benchmark of `r formatC(bf_f, format="f", digits=3)`, i.e. marginally more optimistic than the pre-2020 data alone implied. For UK males the same rounds imply `r formatC(uk_get("male", 2020), format="f", digits=3)`, `r formatC(uk_get("male", 2022), format="f", digits=3)` and `r formatC(uk_get("male", 2024), format="f", digits=3)`, straddling the benchmark of `r formatC(bf_m, format="f", digits=3)` (the 2020 round slightly above, the 2022 and 2024 rounds slightly below). In short, the newest ONS rounds are broadly in line with — not markedly more pessimistic than — the pre-pandemic Bayes-factor estimate; the sharp downward revision happened over the 2012–2018 rounds, and has since levelled off.

The remaining analysis expresses each ONS round and the pre-pandemic Bayes-factor benchmark as a proportion of the mean UK improvement from 1980 to 2010.

```{r}
#| label: mean-longterm-improvement
#| warning: false
mean_lt_imp_uk <-
  dta_e0 %>%
  filter(population == "United Kingdom") %>%
  filter(between(year, 1980, 2011)) %>%
  group_by(sex) %>%
  arrange(year) %>%
  mutate(ch_e0 = e0 - lag(e0)) %>%
  summarise(
    mean_ch_e0 = mean(ch_e0, na.rm = TRUE),
    var_ch_e0 = var(ch_e0, na.rm = TRUE),
    .groups = "drop"
  )

mean_lt_imp_uk
```

```{r}
#| label: tbl-est-as-pct-of-lt
#| warning: false
#| message: false
lt_lookup <- mean_lt_imp_uk %>% select(sex, lt_ch_e0 = mean_ch_e0)

est_as_pct_of_lt <-
  combined_long %>%
  filter(population == "United Kingdom") %>%
  select(round, sex, ONS = ons_mu, `BF (pre-2020)` = bf_mu) %>%
  pivot_longer(c(ONS, `BF (pre-2020)`), names_to = "method", values_to = "ch_e0") %>%
  left_join(lt_lookup, by = "sex") %>%
  mutate(new_as_pct = ch_e0 / lt_ch_e0)

est_as_pct_of_lt %>%
  select(round, sex, method, new_as_pct) %>%
  unite(method, sex, col = "method_sex") %>%
  pivot_wider(names_from = method_sex, values_from = new_as_pct) %>%
  arrange(round) %>%
  kable(digits = 3) %>%
  kable_styling()
```

And as a graph:

```{r}
#| label: fig-est-as-pct-of-lt
#| warning: false
est_as_pct_of_lt %>%
  ggplot(aes(x = round, y = new_as_pct, group = method, shape = method, colour = method)) +
  geom_point() + geom_line() +
  facet_wrap(~sex) +
  expand_limits(y = 0) +
  geom_hline(yintercept = 1) +
  labs(x = "ONS projection round",
       y = "Implied improvement as proportion of 1980-2010 UK trend")
```

Two things stand out. First, from 2011 onwards both the ONS and the pre-pandemic Bayes-factor estimate sat well below the 1980–2010 long-run trend (the horizontal line at 1), confirming that the stalling was real and that the ONS, and the experts advising it, believed the long-run improvement rate was not sustainable. Second, whereas the original (2012–2018) analysis showed each biennial ONS round revising downwards and, for females, still potentially over-optimistic relative to the accumulating data, the extended series shows the ONS rounds levelling off close to the pre-pandemic Bayes-factor benchmark. With the corrected likelihood the evidence for a slowdown is in fact stronger for males than females (see the audit appendix), so the earlier suggestion that the ONS remained too optimistic *mainly* for females does not survive the update. An advantage of the Bayes-factor approach remains that it is trivial to re-run with each new annual data release, updating our beliefs about the underlying improvement rate within minutes of publication.
